Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-25 paper-new/categorical-model.tex
strict and lax homomorphisms of CBPV models

  This means we have 
  \[ \alpha : {A_1 \times A_2} \arr B \cong A_2 \arr (A_1 \arr B) \]
  and
  \[ i : 1 \arr B \cong B \]
  satisfying coherence isomorphisms.
\item A functor $U : \mathcal E \to \mathcal V$ that ``preserves
  powering'' in that every $U(A \arr B)$ is an exponential of $UB$ by $A$
  and that $U\alpha$ and $Ui$ are mapped to the canonical isomorphisms
  for exponentials.
\item A left adjoint $F \dashv U$
\item Distributive finite coproducts in $\mathcal V$
\end{enumerate}

\begin{example}
  Given a strong monad $T$ on a bicartesian closed category $\mathcal
  V$, we can extend this to a CBPV model by defining $\mathcal E$ to
  be the category $\mathcal V^T$ of algebras
\end{example}

% morphisms of CBPV models
Given call-by-push-value models
$\mathcal M_1 = (\mathcal V_1, \mathcal E_1, \arr_1, U_1, F_1)$ and
$\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
A morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$ such that

\begin{enumerate}
  \item $G_{\mathcal{E}} \circ F_1 = F_2 \circ G_{\mathcal{V}}$
  \item $G_{\mathcal{V}} \circ U_1 = U_2 \circ G_{\mathcal{E}}$
  \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
\end{enumerate}

% TODO what about the product and coproducts in V?
% Do we need that G_V(A \times_1 A') = G_V(A) \times_2 G_V(A')
% and likewise for coproducts?

With these definitions, we can define a category whose objects are CBPV
models, and whose arrows are CBPV morphisms.
This category will serve as the setting in which we formulate the definition
of a model of extensional gradual typing.

>>>>

  This means we have 
  \[ \alpha : {A_1 \times A_2} \arr B \cong A_2 \arr (A_1 \arr B) \]
  and
  \[ i : 1 \arr B \cong B \]
  satisfying coherence isomorphisms.
\item A functor $U : \mathcal E \to \mathcal V$ that ``preserves
  powering'' in that every $U(A \arr B)$ is an exponential of $UB$ by $A$
  and that $U\alpha$ and $Ui$ are mapped to the canonical isomorphisms
  for exponentials.
\item A left adjoint $F \dashv U$
\end{enumerate}

\begin{example}
  Given a strong monad $T$ on a bicartesian closed category $\mathcal
  V$, we can extend this to a CBPV model by defining $\mathcal E$ to
  be the category $\mathcal V^T$ of algebras, defining $A \to B$ as
  the powering of algebras, $F$ as the free algebra and $U$ as the
  underlying object functor.
\end{example}

There are two relevant notions of \emph{morphism} of CBPV models:
\emph{strict} and \emph{lax}.
% morphisms of CBPV models
Given call-by-push-value models
$\mathcal M_1 = (\mathcal V_1, \mathcal E_1, \arr_1, U_1, F_1)$ and
$\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
A \emph{strict} morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$
that strictly presere the constructors:
\begin{enumerate}
  \item $G_{\mathcal{E}} \circ F_1 = F_2 \circ G_{\mathcal{V}}$
  \item $G_{\mathcal{V}} \circ U_1 = U_2 \circ G_{\mathcal{E}}$
  \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) = G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
  \item $G_{\mathcal{V}}(1_1) = 1_2$
\end{enumerate}
As well as strictly preserving the corresponding universal morphisms
and coherence isomorphisms.

A lax morphism instead preserves these only up to transformation
\begin{enumerate}
  \item $G_{\mathcal{E}} \circ F_1 \Rightarrow F_2 \circ G_{\mathcal{V}}$
  \item $G_{\mathcal{V}} \circ U_1 \Rightarrow U_2 \circ G_{\mathcal{E}}$
  \item $G_{\mathcal{E}}(A \arr_1 B) \Rightarrow G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) \Rightarrow G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
  \item $G_{\mathcal{V}}(1_1) \Rightarrow 1_2$
\end{enumerate}
Additionally a lax morphism should have a relationship between these
transformations and the universal morphisms, but we will only consider
lax morphisms of thin categories, where such conditions hold
trivially.
% TODO what about the product and coproducts in V?
% Do we need that G_V(A \times_1 A') = G_V(A) \times_2 G_V(A')
% and likewise for coproducts?

With these definitions, we can define categories $\textrm{CBPV}_s$ and
$\textrm{CBPV}_l$ whose objects are CBPV models, and whose arrows are
CBPV morphisms.  This category will serve as the setting in which we
formulate the definition of a model of extensional gradual typing.
